A Process for the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE)

### 2024-11-28 manuscript/manuscript.qmd
advanced with trial selectioin process

library(here)
library(dplyr)
library(knitr)
library(captioner)
library(gt)
library(readxl)

@@

Randomized controlled trials (RCTs) have been the gold standard for establishing the efficacy and safety of medical products. With the advent of the the 21^st^ Century Cures Act directive, the Food and Drug Administration (FDA) established a framework to increasingly include and consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] With 21% of all drug approvals, oncology was the disease area with the most FDA drug approvals in 2023[@senior2024fresh], and especially in the field of precision oncology, RWE has a large potential to complement evidence coming from RCTs. Examples comprise the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible or the use of real-world data (RWD) for biomarker discovery and label extensions among pan-tumor populations that harbor specific genomic and immunological signatures.

However, the validity and transportability of results derived betwen RWE studies and RCTs can depend on many factors and limitations include missing data, small sample sizes, data discontinuity [@lin2018identifying; @joshua2022longitudinal], rapid changes in temporal prescribing patterns and the inability to measure and emulate common eligibility criteria and prognostic factors.[@rider2024emulations] While there are examples of emulations of oncology trials[@rider2024emulations; @merola2023aetion; @merola2024calibrating], a systematic and scaled approach to emulate a diverse set of different oncology trials is necessary to gain confidence the accuracy of RWE studies and to provide an answer as to which questions can be validly answered with which non-interventional study designs and analysis methods, given the data that is available.

@@

The *Emulation of Comparative Oncology Trials with Real-world Evidence* (**ENCORE**) project aims to extend this work to the field of oncology which comes with its own unique challenges that are not necessarily comparable with learnings from other disease areas and which must be systematically explored and understood.

@@

**Systematic process for understanding the validity of RWE for oncology submissions**

@@

```{r}
#| label: tbl-criteria
#| tbl-cap: "Criteria ."
#| tbl-cap-location: top

eligibility_criteria <- read_excel(
  path = here("tables", "helper_tables.xlsx"),
  sheet = "ctgov_criteria_defintions"
  )

eligibility_criteria |> 
  select(-Link) |> 
  filter(Criteria != "Study phase") |> 
  gt() #|> 
  # tab_style(
  #   style = cell_text(weight = "bold"),
  #   locations = 
  #   )
```

@@

rct_selection |> 
  group_by(cancer) |> 
  gt() |> 
  cols_label(
    nctid = md("**NCTID**"),
    acronym = md("**Acronym**"),
    setting = md("**Clinical setting**"),
    line = md("**Line of therapy**"),
    exposures = md("**Treatment comparison**")
    ) |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
    )
```

@@

# copy figure to manuscript
# file_copy(
#   path = path, 
#   new_path = here("manuscript", "Figure_2_attrition.png"), 
#   overwrite = TRUE
#   )
```

[View figure in higher resolution](https://github.com/janickweberpals/encore-process-manuscript/blob/main/figures/process.png)
:::

>>>>

library(here)
library(dplyr)
library(knitr)
library(captioner)
library(gt)
library(readxl)
library(fs)

@@

Randomized controlled trials (RCTs) have been the gold standard for establishing the efficacy and safety of medical products. With the advent of the the 21^st^ Century Cures Act directive[@RWEFDA], the Food and Drug Administration (FDA) established a framework to increasingly include and consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] With 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023,[@senior2024fresh] and especially in the field of precision oncology, RWE has a large potential to complement evidence coming from RCTs. Potential use cases comprise the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible or the use of real-world data (RWD) for biomarker discovery and label extensions among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.

However, the validity and transportability of results derived betwen RWE studies and RCTs can depend on many factors and frequently referenced limitations include missing data, small sample sizes, data discontinuity [@Merola2022; @joshua2022longitudinal], rapid changes in guideline treatment patterns and the inability to measure and emulate common eligibility criteria and prognostic factors in RWD.[@rider2024emulations] While there are already published examples of oncology trial emulations[@rider2024emulations; @merola2023aetion; @merola2024calibrating], a systematic and scaled approach to emulate a diverse set of different oncology trials in various heterogeneous databases is necessary to gain confidence in the accuracy of RWE studies and to provide an answer as to which questions can be validly answered with which non-interventional study designs and analysis methods.

@@

The *Emulation of Comparative Oncology Trials with Real-world Evidence* (**ENCORE**) project [@encoreFDA] aims to extend this work to the field of oncology which comes with its own unique set of challenges which must be systematically explored and understood. Building on a process co-developed with the FDA through RCT DUPLICATE[@wang2023emulation], this expansion to oncology is going to emulate 12 randomized clinical trials using multiple EHR data sources. The process includes an emphasis on transparency with documented assessment of data fitness of the RWD source for each trial[@rivera2024oncology; @gatto2022structured] and conducting extensive sensitivity analyses to assess robustness of findings and trial eligibility criteria.

The objectives of this project are to develop state-of-the-art methodological approaches and apply these to create insights that may provide guidance on the potential use of RWE for regulatory science in oncology. This includes the systematic evaluation of the suitability of data in relation to the study design and statistical analysis by emulating 12 oncology trials across four cancers and assessing the agreement of treatment effect estimates between RCTs and their respective emulations.

In this process paper, we describe the design and process for the selection of the 12 oncology RCTs, the assessment of the database quality and selection, protocol development, study design and statistical analysis and final agreement metrics to evaluate the concordance between RCTs and emulations.

@@

A visual summary of the entire systematic process from trial selection to final results is provided in @fig-process.

## Trial selection

The focus of ENCORE is to maximize potential learnings on when RWE studies can or cannot yield similar results compared to RCTs. To that end, the focus of the project is on trials of therapies for the most common cancers and/or cancers for which there has been substantial therapeutic development in recent years. After careful review exchange with clinical and regulatory experts, four cancer indications were identified including lung cancer, breast cancer, colorectal cancer and multiple myeloma. For each cancer we aim to conduct three trial emulations which will be implemented using multiple databases which will be accessible for the scope of this project (i.e. the total number of emulations will equal 12 trials x n databases which are found fit-for-purpose for each trial).

The trial selection will follow a semi-automated process for which we will document the eligibility criteria and decisions, resulting in a CONSORT diagram showing reasons for excluding RCTs. The search will be conducted using the AACT database which is s a publicly available relational database developed and maintained by the Clinical Trials Transformation Initiative (CTTI) which contains all information (protocol and result data elements) about every study registered in ClinicalTrials.gov.[@tasneem2012database] To identify eligible trials, we will use a combined search query strategy of the National Library of Medicine (NLM)-controlled *MeSH* term and a free keyword search for the respective cancer indication in the *conditions*, *studies* and *detailed_descriptions* fields of each trial entry on ClinicalTrials.gov.

Eligible trials need to fulfill the following basic criteria:

-   Interventional

-   Randomized

-   Intervention model: parallel assignment

-   Industry-sponsored

-   Trial start in 2011 or later

-   Primary purpose was to study treatment effects

-   Primary or secondary endpoint must be overall survival

-   Recruitment status: ‘Completed’ or ‘Active, not recruiting’

-   Feasibility and clinical relevance

The rationale and operationalization of each criterion is listed in detail in @tbl-criteria. While all but the last criteria are highly objective and can be used to filter potential trials in an automated fashion, the last criterion on emulation feasibility and clinical relevance involves extensive human review, a thorough feasibility assessment in context of available databases and discussion with clinical and regulatory experts. The aim is to identify a broad and diverse set of potentially feasible RCTs for which there is reasonable believe that key study parameters can be emulated and there is a high enough clinical relevance for regulatory decision-making (e.g., paradigm-changing trials).

## Databases

## Protocol development

## Application of the target trial emulation framework to study design and statistical analysis

Agreement metrics

@@

```{r}
#| label: tbl-criteria
#| tbl-cap: "Criteria to select eligible trials for emulation in ENCORE."
#| tbl-cap-location: top

# read excel blueprint
eligibility_criteria <- read_excel(
  path = here("tables", "helper_tables.xlsx"),
  sheet = "ctgov_criteria_defintions"
  )

# convert to gt table
tbl_eligibility_criteria <- eligibility_criteria |> 
  select(-Link) |> 
  filter(Criteria != "Study phase") |> 
  gt() |> 
  tab_style(
    style = cell_text(weight = "bold"),
      locations = cells_column_labels()
    )

# save table as .docx document
tbl_eligibility_criteria |> 
   gtsave(
    filename = here("manuscript", "Table_1_trial_eligibility.docx")
    )

# save image to display within manuscript document
tbl_eligibility_criteria |> 
  gtsave(
    filename = here("tables", "Table_1_trial_eligibility.png"),
    expand = 10, # amount of whitespace,
    zoom = 5 # higher resolution
    )

include_graphics(path = here("tables", "Table_1_trial_eligibility.png"))
```

@@

tbl_rct_selection <- rct_selection |> 
  group_by(cancer) |> 
  gt() |> 
  cols_label(
    nctid = md("**NCTID**"),
    acronym = md("**Acronym**"),
    setting = md("**Clinical setting**"),
    line = md("**Line of therapy**"),
    exposures = md("**Treatment comparison**")
    ) |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
    )

# save table as .docx document
tbl_rct_selection |> 
   gtsave(
    filename = here("manuscript", "Table_2_trial_selection.docx")
    )

# save image to display within manuscript document
tbl_rct_selection |> 
  gtsave(
    filename = here("tables", "Table_2_trial_selection.png"),
    expand = 10, # amount of whitespace,
    zoom = 5 # higher resolution
    )

include_graphics(path = here("tables", "Table_2_trial_selection.png"))
```

@@

# copy figure to manuscript
file_copy(
  path = here("figures", "process.png"),
  new_path = here("manuscript", "Figure_1_process.png"),
  overwrite = TRUE
  )
```

[View figure in higher resolution here](https://github.com/janickweberpals/encore-process-manuscript/blob/main/figures/process.png)
:::
